---
title: "Commentary on Lab 12 - Smoking during pregnancy"
author: "Benjamin Egan"
output: github_document
---

#### Created 4/14/2025

Link to the assignment page, at the time of creation: https://web.archive.org/web/20250414135227/https://datascience4psych.github.io/DataScience4Psych/lab12.html

Link to assignment page as of `r format(Sys.time(), '%B %d %Y')`: https://datascience4psych.github.io/DataScience4Psych/lab12.html

## Commentary on Lab 12

I will be walking through lab 12, designed to practice data manipulation, visualization, hypothesis testing, and calculating confidence intervals. Questions and comments from the lab that I will be addressing will be in bold. My commentary will always have "Commentary:" infront. This can make it easier to quickly skim through the document by searching for the word commentary.

```{r load-packages, message=FALSE, warning = FALSE, echo = FALSE}
library(tidyverse) 
library(tidymodels)
library(openintro)
seed <- set.seed(123)
```

Loading the data
```{r load data}
data(ncbirths)
#view(ncbirths)
```

#### Before analyzing any new dataset, it’s important to get to know your data. Start by summarizing the variables and determining if their data type. Are they categorical? Are they numerical? For numerical variables, check for outliers. If you aren’t sure or want to take a closer look at the data, create a graph.

```{r examine data}
summary(ncbirths)
```

By using summary(), I know there are 7 categorical variables and 6 numerical variables.

I way overengineered looking for outliers, getting help from Chat GPT.
```{r outlier check, warning = FALSE, message = FALSE}


base_outlier_check <- function(variable) {
  ncbirths %>%
  ggplot(aes(x = {{ variable }})) +
    geom_histogram(stat = "count") +
    labs(title = rlang::as_label(rlang::enquo(variable)), y = NULL)
}

map(c("fage", "mage","weeks","visits","gained","weight"), ~ base_outlier_check(!!sym(.x)))

```

mage could have an outlier on the upper end. I don't think weeks has an outlier, even though it's skewed left. visits may have an outlier on the upper end. weight might have one or two on the upper end, but it's skewed left.

Commentary: This was pretty basic, and I made it way harder than it had to be. This could, however, be a good moment to throw it back to lab 8 and show a way to use a function. I would just guide them on the labeling bit, as it can be tricky. You can use do what I did and use base R. This allows you to input all the variables at once using map(). If you use tidyverse for the label and use deparse(substitute(variable)), you have to individually input each function and cannot mass input.

#### What are the cases in this data set? How many cases are there in our sample?

```{r number cases}

ncbirths %>%
nrow()

```

There are 1000 cases, as each row represents an individual birth.

### Create a filtered data frame called ncbirths_white that contains data only from White mothers. Then, calculate the mean of the weights of their babies.

```{r filter only white}

ncbirths_white <- ncbirths %>%
  filter(whitemom == "white")

mean(ncbirths_white$weight)


```

The mean weight of the white babies is 7.25 lbs. The hypothesis states that there is no change in weight since 1995 (i.e. mu = 7.43). Since white babies weigh less, we need to figure out if this is a meaningful difference.

Commentary: You/someone else included a note in the code that says "Mason make this clearer to why we need to actually do this." I don't think you need to explain, if they watched the videos or viewed the slides. What I would like more clarification on is question 3. My imagination for an answer is to create a visual for the data and see if mu falls within the CI in this sample.

### Are the criteria necessary for conducting simulation-based inference satisfied? Explain your reasoning.

```{r should we baby weight hypothesis test}

ncbirths_white %>%
  ggplot(aes(
    x = weight
  ))+
  geom_histogram(stat = "count", color = "blue")+
  geom_vline(xintercept = 7.43, 
               color = "white", 
               linetype = "dashed")+
   geom_vline(xintercept = 7.25, 
               color = "red", 
               linetype = "dashed")+
  annotate("text", 
           x = 4, y = Inf, 
           label = "White = Mu, Red = Sample mean", 
           vjust = 2, 
           color = "black")+
  theme_bw()+
  labs(
    x = "Weight in lbs",
    y = NULL,
    title = "Distriubtion of weight",
    subtitle = "How does this shape up to the popluation mean?"
  )

```
Based on this graph, I would guess that we probably would fail to reject, considering the mean weight is so close to the population weight. But this is only one sample, so it makes sense to use bootstrapping to better estimate the difference.

## Steps for hypothesis testing
- take a bootstrap sample of from the original sample
- calculate this bootstrap sample’s mean
- repeat these two steps a large number of times to create a bootstrap distribution of means centered at the observed sample mean,
- shift this distribution to be centered at the null value by subtracting / adding X to all bootstrap mean (X = difference between mean of bootstrap distribution and null value)
- calculate the p-value as the proportion of bootstrap samples that yielded a sample mean at least as extreme as the observed sample mean.

```{r bootstrap weight}
seed

white_weight_bootstrap <- ncbirths_white %>%
  specify(response = weight) %>% 
  generate(reps = 15000, type = "bootstrap") %>% 
  calculate(stat = "mean")

mean_white_weight_boostrap <- mean(white_weight_bootstrap$stat)

white_weight_bootstrap <-  white_weight_bootstrap %>%
  mutate(Cstat = (stat - 7.43))

white_weight_bootstrap %>%
  summarize(lower = quantile(Cstat, 0.025),
            upper = quantile(Cstat, 0.975))


white_weight_bootstrap %>%
  ggplot(mapping = aes(x = Cstat)) +
  geom_histogram() +
  labs(title = "Bootstrap distribution of means for white baby weight", y = NULL, x = "Means")+
  geom_vline(xintercept = 0, 
               color = "black", 
               linetype = "solid")+
  annotate("text", 
           x = -.03, y = Inf, 
           label = "Pop mean", 
           vjust = 2, 
           color = "black")+
  geom_vline(xintercept = -0.2847633,                 #Lower CI
               color = "blue", 
               linetype = "dashed")+
  geom_vline(xintercept = -0.07682038,               #Upper CI
               color = "blue", 
               linetype = "dashed")+
  annotate("text", 
           x = -0.31, y = Inf, 
           label = "95% CI", 
           vjust = 2, 
           color = "black")+
  theme_bw()
  

# p value
pvalue_white_weight <- white_weight_bootstrap %>%
  mutate(sig = case_when(
    stat <= mean_white_weight_boostrap ~ 1,
    stat >= mean_white_weight_boostrap ~ 0,
  ))


white_weight_sig <- pvalue_white_weight %>%
  filter(sig == 1)

count(white_weight_sig)/15000

```


95% CI:(7.145237, 7.35318)

Based on this, it's clear that the population mean falls outside the confidence interval. We would reject the null and say that babies' weight has decreased since 1995.

To calculate the p value, I created a variable that designated whether the mean in the bootstrap was lower than the observed mean of 7.25. I then divided the number of significant ones by the total number of observations in the bootstrap to get a p of .50, indicating that the sample means should not be significantly different from the popluation mean.

Commentary: I would break out the step by step instructions. It was easy to figure out how to make the graph based on the slides, so maybe add in a tip to go back and reference the slides? I will say that I believe I messed this up somewhere along the way, as my p value should be reflective of the fact that the population mean of 7.43 is way outside the confidence interval I estimated. If I had to guess, I either messed up the calculation of the p value or my interpretation of the p value.


## Baby Weight vs Smoking

#### Make side-by-side box plots displaying the relationship between habit and weight. What does the plot highlight about the relationship between these two variables?

Commentary: I would specify here if you want us to continue to utilize only White babies or all babies. I am assuming you want us to use all babies here.

```{r box plot habit weight}

ncbirths %>%
  ggplot(aes(
    x = habit,
    y = weight
  ))+
    geom_boxplot()+
  labs(
    x = "Smoking Status",
    y = "Weight in lbs",
    title = "Relationship between smoking and weight"
  )

```

It appears that there are a lot of underweight babies for non-smokers compared to smokers. This could be due to sample differences, as non-smokers could have more babies overall.

#### Before continuing, create a cleaned version of the dataset by removing any rows with missing values for habit or weight. Name this version ncbirths_clean.

```{r clean smoking data}

ncbirths_clean <- ncbirths %>%
  filter(!habit == "NA")

ncbirths_clean %>%
  ggplot(aes(
    x = habit,
    y = weight
  ))+
    geom_boxplot()+
  labs(
    x = "Smoking Status",
    y = "Weight in lbs",
    title = "Relationship between smoking and weight"
  )

```

We can see that we cleaned the data to filter out the person who didn't answer the smoking question.

#### Calculate the observed difference in means between the baby weights of smoking and non-smoking mothers.

```{r comapre mean weight smokers}

ncbirths_clean %>%
  group_by(habit) %>%
  summarize(mean_weight = mean(weight))

```

We can see that non-smoking babies have a higher mean weight that smoking babies.


#### Write the hypotheses for testing if the average weights of babies born to smoking and non-smoking mothers are different.

H0: Mu1 - Mu2 = 0 OR Mu1 = Mu2 (No difference between smokers and non-smokers on baby weight)

H1: Mu1 - Mu2 ≠ 0 OR Mu1 ≠ Mu2 (There is a difference between smokers and non-smokers on baby weigh)

Commentary: I TA for the 312 class and have hypothesis testing essentially tattooed on my frontal lobe.

#### Construct a 95% confidence interval for the difference between the average weights of babies born to smoking and non-smoking mothers.

*I did them out of order, this is Q10

```{r CI for smoking}
view(ncbirths_clean)

ci_vals <- ncbirths_clean %>%
  group_by(habit) %>%
  summarize(
    mean_weight = mean(weight),
    se = sd(weight) / sqrt(sum(!is.na(weight))),
    lower = mean_weight - 1.96 * se,
    upper = mean_weight + 1.96 * se
  )
ci_vals
```
CI for non-smokers (7.043530, 7.245016)
CI for smokers (6.586688, 7.070772)

```{r visual CI for smoking}

ncbirths_clean %>%
ggplot(aes(
  x = weight, 
  fill = habit, 
  color = habit)) +
  geom_density(alpha = 0.4) +
  
  # Add CI
  geom_vline(data = ci_vals, 
             aes(xintercept = lower, 
                 color = habit), 
             linetype = "dashed") +
  geom_vline(data = ci_vals, 
             aes(xintercept = upper, 
                 color = habit), 
             linetype = "dashed") +
  scale_color_manual(
    name = "Smoking Status",
    values = c("nonsmoker" = "blue", "smoker" = "red"))+
  scale_fill_manual(
    name = "Smoking Status",
    values = c("nonsmoker" = "skyblue", "smoker" = "pink"))+
  
  #I got some help from ChatGPT to make it better. I clearly could've used a string and wrote out the CI, but this looked cooler
  annotate("text", 
         x = 2, y = 0.33, hjust = 0, size = 3.5,
         label = paste0("Non-smoker 95% CI: [", 
                        round(ci_vals$lower[ci_vals$habit == "nonsmoker"], 2), ", ", 
                        round(ci_vals$upper[ci_vals$habit == "nonsmoker"], 2), "]"),
         color = "blue") +
annotate("text", 
         x = 2, y = 0.30, hjust = 0, size = 3.5,
         label = paste0("Smoker 95% CI: [", 
                        round(ci_vals$lower[ci_vals$habit == "smoker"], 2), ", ", 
                        round(ci_vals$upper[ci_vals$habit == "smoker"], 2), "]"),
         color = "red")+
  
  labs(
    x = "Baby weight in lbs",
    y = NULL,
    title = "Baby weight distribution based on smoking"
  ) +
  theme_bw()

```

Based on this graph, it looks like the two distributions heavily overlap, and there appears to be some basic overlap for the CIs.

#### Run the appropriate hypothesis test, calculate the p-value, and interpret the results in context of the data and the hypothesis test.

##### Commentary 
I will admit that I was completely lost on how to get the p value. I looked back through the class notes and here is what I came up with:

Idea 1: Create the null distribution <br/>
I was thinking about somehow creating the null distribution that assumes the mean difference is zero. I could possibly do this by somehow bootstrapping the weight data but telling the simluation that the resulting outcome should be two stats, one for smokers and one for non-smokers

Idea 2: two boostraps and join <br/>
For this, I would filter based on smoking status. I would then bootstrap for each of the two datasets, and then join them together. The downside to this is that I don't know if I would need to bootstrap the combined bootstraps to truly achieve what I wanted to do.

Idea 3: cry <br/>
No not actually, but I am thinking that I could somehow create difference scores and bootstrap the difference scores. I only think this would work if there is equal sample size, which I know for a fact there isn't based on the summary() look at the variables.

I'm thinking that I'll need to use idea one, since I'm not confident idea 2 would give me the best outcomes.

```{r p value for smoking}
seed

smoking_pvalue_test <- ncbirths_clean %>%
  group_by(habit) %>%
  specify(response = weight) %>% 
  generate(reps = 15000, type = "bootstrap") %>% 
  calculate(stat = "mean")

view(smoking_pvalue_test)

```

Ok so I fought with chat heavily to figure out how to do it this way via idea 1, and I have zero clue why group_by isn't doing it's job. So now I'll try idea two, but using chat's help

```{r p value for smoking attempt 2}
# Load necessary libraries
library(dplyr)
library(infer)


# Split the dataset by habit (smoker vs nonsmoker)
smoker_data <- ncbirths_clean %>% filter(habit == "smoker")
nonsmoker_data <- ncbirths_clean %>% filter(habit == "nonsmoker")

# Perform bootstrap for smokers
smoker_bootstrap <- specify(smoker_data, response = weight) %>%
  generate(reps = 15000, type = "bootstrap") %>%
  calculate(stat = "mean")

# Perform bootstrap for non-smokers
nonsmoker_bootstrap <- specify(nonsmoker_data, response = weight) %>%
  generate(reps = 15000, type = "bootstrap") %>%
  calculate(stat = "mean")

# Combine both bootstrap results into one data frame
white_weight_habit_bootstrap <- bind_rows(
  smoker_bootstrap %>% mutate(habit = "smoker"),
  nonsmoker_bootstrap %>% mutate(habit = "nonsmoker")
)

```


I went back and asked it to do it only using the libraries required thus far.

```{r p value for smoking attempt 3}

# Split the dataset by 'habit' (smoker vs nonsmoker)
smoker_data <- ncbirths_clean %>% filter(habit == "smoker")
nonsmoker_data <- ncbirths_clean %>% filter(habit == "nonsmoker")

# Define a function to perform bootstrapping
bootstrap_mean <- function(data, reps = 15000) {
  # Create bootstrap samples and calculate means
  replicate(reps, {
    sample_data <- sample(data$weight, replace = TRUE)
    mean(sample_data)
  })
}

# Perform bootstrap for smokers
smoker_bootstrap_means <- bootstrap_mean(smoker_data, 15000)

# Perform bootstrap for non-smokers
nonsmoker_bootstrap_means <- bootstrap_mean(nonsmoker_data, 15000)

# Combine both bootstrap results into one data frame
white_weight_habit_bootstrap <- tibble(
  habit = c(rep("smoker", 15000), rep("nonsmoker", 15000)),
  bootstrap_mean = c(smoker_bootstrap_means, nonsmoker_bootstrap_means)
)

```


Another attempt using ChatGPT. Here is what I plugged into chat

What if i try creating a loop using for or if or while. The loop specifies that I filter for only smoker data, simulate 1500, and apply a categorical variable with the designation "smoker". I then repeat the sequence, except I now filter for only nonsmoker data pply a categorical variable with the designation "nonsmoker" 

It gave me a response I didn't like, and asked "Is it possible to restructure or recreate the code such that we won't have to combine the boostraps and that it will do it all in one go?"

```{r p value for smoking attempt 4}

# Set habit groups and number of reps
habits <- c("smoker", "nonsmoker")
n_reps <- 15000

# Initialize an empty tibble to collect all bootstrap results
white_weight_habit_bootstrap <- tibble()

# Loop through both habit groups and perform bootstrap
for (h in habits) {
  
  # Filter the data for the current group
  current_data <- ncbirths_clean %>% filter(habit == h)
  
  # Perform bootstrap in one go (faster than nested loop)
  boot_means <- map_dbl(1:n_reps, ~ mean(sample(current_data$weight, replace = TRUE)))
  
  # Append results directly to the main tibble
  white_weight_habit_bootstrap <- bind_rows(
    white_weight_habit_bootstrap,
    tibble(habit = h, bootstrap_mean = boot_means)
  )
}

# View result
view(white_weight_habit_bootstrap)


```


Commentary: To figure out what is happening with group_by, I simulated data with extreme means.

```{r simluate data}
seed

small <- rnorm(100, mean = 10, sd = 5)

dfsmall <- data.frame(
    score = small,
    group = 1
  )

large <- rnorm(100, mean = 1000, sd = 5)

dflarge <- data.frame(
    score = large,
    group = 2
  )

simulation <- full_join(dfsmall,dflarge)



simluation_test1 <- simulation %>%
  group_by(group) %>%
  specify(response = score) %>% 
  generate(reps = 15000, type = "bootstrap") %>%
  calculate(stat = "mean") %>%
  ungroup()

view(simluation_test)

```


Commentary: turns out it just ignores the group_by, as the bootstrapping provides means that encompass the entire data.


Ok after talking to 4/16 Mason, we figured out she intended for us to run a t-test.

```{r t test for smoking}

t.test(weight ~ habit, data = ncbirths_clean)

```

